Paper Proposal for American Political Development (POLSCI 688)

### 2019-12-17 paper/paper-draft.tex
premption plot

\title{Paper Proposal for POLSCI 688}
\author{Ben Goehring}
\date{\today}

@@

\subsection*{City-state relations}
State interference in municipal governance dates back to the founding of the United States. Following the passage of the Constitution, which does not mention cities or other types of sub-state municipalities, state legislatures began to exert control over cities chartered in the colonial era in order to reduce local elites' authority. While state intervention was relatively rare in the first half of the 19th Century, states took a much more active role in municipal affairs in the latter half of the century. The challenges wrought by urbanization and industrialization led many reformers and state legislators to see the states as having a necessary role in governing large cities while concerns over urban corruption and nativist prejudices further galvanized efforts to increase state control over city affairs (\cites[p. 53-57]{bermanLocalGovernmentStates2003}[p. 9]{kraneHomeRuleAmerica2000}).

@@

\subsection*{History of preemption}
The emergence of legislative Home Rule in the U.S. states set the stage for the preemption battles of recent decades.\footnote{need info about how many states have Home Rule} Under such a legal regime, cities enjoy expanded authority but face the threat of interference from either state courts or legislatures \parencite{dillerIntrastatePreemption2007,goodmanStateLegislativeIdeology2019}. Court-driven preemption occurs when an interest group, the state, or another plaintiff appeals to a state court that a local law surpasses the city's policymaking authority. While this court-centered form of preemption is important and constitutes a significant share of overall preemption activity \parencite{swansonStateGovernmentPreemption2018,dillerIntrastatePreemption2007}, my focus here is on preemption stemming from the actions of state legislatures. Legislative preemption occurs when a state steps in to explicitly ban local policymaking on a certain topic, such as plastic bags or guns, effectively setting a statewide ceiling on regulation.\footnote{States have also considered but not passed even wider forms of preemption that ban policymaking across expansive policy domains, such as private property, business, and commerce \parencite[p. 2007]{briffaultChallengeNewPreemption2018}.}

@@

In many states, the tobacco industry succeeded in advocating for laws preempting local tobacco regulations. Between 1982 and 1996, 31 states enacted approximately 120 different tobacco-related preemption laws \parencite{centersfordiseasecontrolPreemptiveStateTobaccoControl1999}. Although the number of laws preempting tobacco regulation in certain domains such as indoor smoking in public and private buildings decreased over the following fifteen years \cite{centersfordiseasecontrolStatePreemptionLocal2010}, 13 states still preempt at least some local smoke free ordinances. In four of those states--Connecticut, Florida, New Hampshire, and South Dakota-- all local smoke-free ordinances are preempted \parencite{centersfordiseasecontrolStatePreemptionLocal2010}.\footnote{Some of these states' preemption provisions resulted from court decisions rather than legislative acts. While separating instances of court and legislative preemption is beyond the scope of this paper proposal, I will differentiate the two once I receive the tobacco data.}

The efforts of the tobacco industry to prevent local regulations were noticed by other interest groups. Working on behalf of firearm enthusiasts and producers, the National Rife Association (NRA) became a key advocate for state laws preempting gun control regulations \parencite{hicksHomeRuleBe2018}. Today, 43 states preempt at least some local regulation of guns and ammunition \parencite{giffordslawcenterPreemptionLocalLawsn.d.,schraggerStatePreemptionLocal2017}, with many of the laws dating back to the 1980s and 1990s. Of the eleven states that do not allow any firearm or ammunition ordinances, for instance, nine of them passed their preemption laws before 2000 \parencite{schraggerStatePreemptionLocal2017}. More recently, some states have adopted gun control preemption laws that impose severe penalties on non-compliant local officials and cities. While such penalties are not exclusively associated with laws preventing local firearm regulation, many of the penalties for not complying with gun preemption laws are especially severe. Kentucky criminalizes not complying with the state preemption law, while local officials in Florida can be fined and even removed from office for not complying the state's gun preemption law \parencite{briffaultChallengeNewPreemption2018}.

@@

@@

The first condition concerns the threat of local policymaking. The legal regimes governing cities' authority empower local policymaking to a greater extent in some states than in others, resulting in a greater likelihood of municipal action. If cities cannot pass regulations in the first place, states should not be expected to take the time to preempt specific municipal ordinances. The second condition concerns the ideological gap between cities and the state. If cities and states possess different preferences over the given regulatory measure, we should expect states to be more willing to preempt local authority. In states where city and state preferences are closer to one another, states will be more willing to let localities engage in policymaking. Finally, the third condition concerns the activity of relevant interest groups. The narrative above suggests that tobacco lobbying firms, the NRA, and ALEC have played a key role in mobilizing states to clamp down on local policymaking. Where these groups are more powerful, we should see more preemption activity.

@@

I hope to show that these conditions should apply to all preemption activity, regardless of the policy area. While the broad applicability of the hypotheses is an advantage, the reality of the data on state preemption and local policymaking means that testing hypotheses across multiple policy domains is a distant goal. Collecting data on cities' policymaking (which is necessary to test whether the realization of the threat of local action spurs states to act preemptively) is no small task. I only know of one resource, the U.S. Tobacco Control Laws Database (TCLD) maintained by the American Nonsmokers' Rights Foundation,\footnote{\url{https://no-smoke.org/materials-services/lists-maps/}} that tracks both local ordinances and state preemption laws. While collecting data on cities' attempts to pass other regulations is important (especially in the policy areas that have seen the sharpest rise in preemption activity in recent years), my aims are more modest. I hope to use the TCLD's data on local tobacco regulation and state preemption over the past 40 years to test whether my three conditions hold. In the following pages, I provide further justification of the three conditions and offer concrete means of testing them in the context of tobacco preemption.

\subsubsection*{Hypothesis 1}
The first condition concerns cities' capacities to pass regulations. As described above, cities have often lacked this power. Early formulations of Home Rule, paired with narrow interpretations of cities' sphere of policymaking authority, hindered cities' ability to legislate as they saw fit. Since mid-century, many states have adopted legislative Home Rule regimes that permit a greater degree of state policymaking while allowing state courts and legislatures to intervene and deem local ordinances in conflict with state law. Today, while legislative Home Rule predominates, cities are governed by a patchwork of state legal regimes, with different implications for the likelihood of state preemption. In states where Dillon's Rule is still frequently applied by judges, cities possess limited policymaking authority. They can only legislate in policy domains that have been explicitly permitted by the state. While states can theoretically grant wide authority to cities and then preempt ordinances by curtailing city's delegated power, states that employ Dillon's Rule usually delegate narrow powers to cities in the first place. As a result, cities have little capacity to legislate and if they do act outside the scope of their delegated authority, it usually falls to the state courts to determine if the ordinance is legal \parencite{dillerIntrastatePreemption2007}.

In other states, cities' authority is structured according to the earlier formulation of Home Rule. Cities enjoy authority over matters of local concern, but the bounds of ``local concern'' is up to state judges' interpretation. Consequently, if a city ordinance is preempted, it is usually the result of a court deciding that the city acted outside of its local area of concern \parencite{dillerIntrastatePreemption2007}. In contrast, legislative Home Rule empowers cities to pass any ordinance that does not conflict with state law. While cities enjoy broad authority in such a regime, state legislatures, as discussed above, also possess the authority to intervene and curtail the bounds of local autonomy. Therefore, it is in states that govern cities with legislative Home Rule that we should expect to see the greatest level of preemption activity.

Unfortunately, given the complexities of state-local relations, there is no clear method for classifying states' legal regimes. Even the distinction between Home Rule and Dillon's Rule has been challenged \parencite{richardsonDillonRuleMars2011}. In a review of different classification schemes, \textcite{dillerIntrastatePreemption2007} notes that some scholars count 48 states with Home Rule while others count 45. Differentiating between the type of Home Rule is even trickier. One study counts 26 states with legislative Home Rule while others argue that the vast majority of Home Rule states employ legislative Home Rule. Moreover, classification disagreements do not only occur at the margin. In their analysis of the causes of state preemption, \textcite{fowlerStatePreemptionLocal2019}, citing \textcite{kraneHomeRuleAmerica2000}, only treat 10 states as using Home Rule.

Classifying the legal regimes governing cities in the states is a challenging but necessary step for measuring how the threat of local action affects states' decisions to preempt local ordinances. While states' legal regimes provide one perspective on the threat of local action, another more concrete measure is whether a city in a state has already passed a tobacco regulation. Unlike the state's legal rules governing cities, which measures the potential for action, whether a city has already regulated the sale or consumption of tobacco captures the realization of the threat. Controlling for cities' policymaking is notably absent from recent quantitative analyses of preemption \parencite{fowlerStatePreemptionLocal2019,goodmanStateLegislativeIdeology2019,flavinExplainingStatePreemption2019}, likely due to the difficulty of collecting data on cities' policies. Luckily, the TCLD systematically tracks local tobacco regulations alongside states' preemption laws. It includes the tobacco regulations of 98\% of the cities in the U.S. with more than 75,000 inhabitants, allowing me to test my hypotheses across all fifty states.

\subsubsection*{Hypothesis 2}
The second condition contends that diverging policy preferences between the state government and at least one city in the state increases the likelihood of state preemption. Struggles between cities and states over the bounds of local autonomy usually pit conservative state governments against liberal cities \parencite{rapoportBlueCitiesBattle2016,riverstone-newellRiseStatePreemption2017}, although this is not always the case.\footnote{For instance, in 2013, Oregon preempted cities from regulating genetically-modified crops and, in 2017, New York overturned New York City's tax on plastic bags \parencite{mckinleyCuomoBlocksNew2017,scharffHyperPreemptionReordering2018}.} Given the increasing number of state preemption laws targeting local economic regulations and increasing polarization between Republicans and Democrats on economic and labor issues over the past two decades \parencite{grumbachBackwatersMajorPolicymakers2018}, Republican control of government and state conservatism, unsurprisingly, play a key explanatory role in recent studies of preemption \parencite{hicksHomeRuleBe2018,riverstone-newellRiseStatePreemption2017,flavinExplainingStatePreemption2019,fowlerStatePreemptionLocal2019}.

While Republican control of government and conservatism certainly correlate with recent preemption activity, we do not yet have a clear idea of whether partisanship and ideology matter in absolute or relative terms. In other words, we do not know whether Republican control of government and conservatism inherently lead to preemption or whether the connection hinges on the relative difference between the preferences and ideology of the state and its cities. Should we expect preemption to be just as likely in North Dakota and Texas, since they are both conservative states controlled by Republicans? Or does the fact that Texas contains much more progressive municipalities mean that it is more likely to preempt local ordinances? Based on the above discussion of preemption's history, where preemption often seems to occur as a response to local actions, the latter framing seems to be a more precise formulation of the role of partisanship and ideology. It is also more generalizable, as it does not assume that certain parties and ideologies always pursue preemption.

While treating ideology and policy preferences in relative terms is important, it is also challenging. Measuring the liberal ideology of cities is difficult, especially over multiple decades. One, albeit imperfect, approach is to use the share of votes earned by Democratic presidential candidates in urban counties. Figure \ref{fig:ideology} uses this approach to display the growing ideological divide between states and cities. The dashed lines in the figure show the increasing relative support Democratic presidential candidates enjoy in the three largest counties in states that support the Democratic and Republican candidate, respectively. Over time, the largest counties have become increasingly more liberal on average than their states. This is especially true for counties in states that support the Republican presidential candidate (red dashed line).

@@

The above discussion suggests that we need to move beyond Republican control of government and conservative ideology as intrinsically linked to state preemption of local laws. While recent decades' increasing polarization and diverging preferences between rural and urban areas means that conservative, Republican-controlled states are often the source of recent preemption activity, it does not seem that this has always been the case. The ideological and partisan geography of the U.S. in the late 1980s and 1990s suggests that preemption must have been occurring in Democratic states. It will take getting the TCLD data in-hand for me to determine whether this circumstantial evidence holds up to further scrutiny.

>>>>

\title{Paper Proposal for American Political Development (POLSCI 688)}
\author{Ben Goehring}
\date{\today}

@@

\subsection*{City-state relations}
State interference in municipal governance dates back to the founding of the United States. Following the passage of the Constitution, which does not mention cities or other types of sub-state municipalities, state legislatures began to exert control over cities chartered in the colonial era in order to reduce local elites' authority. While state intervention was relatively rare in the first half of the 19th Century, states took a much more active role in municipal affairs in the latter half of the century. The challenges wrought by urbanization and industrialization led many reformers and state legislators to see the states as having a necessary role in governing large cities, while concerns over urban corruption and nativist prejudices further galvanized efforts to increase state control over city affairs (\cites[p. 53-57]{bermanLocalGovernmentStates2003}[p. 9]{kraneHomeRuleAmerica2000}).

@@

\subsection*{History of preemption}
The emergence of legislative Home Rule in the U.S. states set the stage for the preemption battles of recent decades. Under such a legal regime, cities enjoy expanded authority but face the threat of interference from either state courts or legislatures \parencite{dillerIntrastatePreemption2007,goodmanStateLegislativeIdeology2019}. Court-driven preemption occurs when an interest group, the state, or another plaintiff appeals to a state court that a local law surpasses the city's policymaking authority. While this court-centered form of preemption is important and constitutes a significant share of overall preemption activity \parencite{swansonStateGovernmentPreemption2018,dillerIntrastatePreemption2007}, my focus here is on preemption stemming from the actions of state legislatures. Legislative preemption occurs when a state steps in to explicitly ban local policymaking on a certain topic, such as plastic bags or guns, effectively setting a statewide ceiling on regulation.\footnote{States have also considered but not passed even wider forms of preemption that ban policymaking across expansive policy domains, such as private property, business, and commerce \parencite[p. 2007]{briffaultChallengeNewPreemption2018}.}

@@

The tobacco industry succeeded in advocating for laws preempting local tobacco regulations in a number of states. Between 1982 and 1996, 31 states enacted approximately 120 different tobacco-related preemption laws \parencite{centersfordiseasecontrolPreemptiveStateTobaccoControl1999}. Although the number of laws preempting tobacco regulation in certain domains such as indoor smoking in public and private buildings decreased over the following fifteen years, 13 states still preempt at least some local smoke free ordinances \parencite{centersfordiseasecontrolStatePreemptionLocal2010}. In four of those states---Connecticut, Florida, New Hampshire, and South Dakota---all local smoke-free ordinances are preempted \parencite{centersfordiseasecontrolStatePreemptionLocal2010}.\footnote{Some of these states' preemption provisions resulted from court decisions rather than legislative acts. While separating instances of court and legislative preemption is beyond the scope of this paper proposal, I will differentiate the two once I receive the tobacco data.}

Other interest groups noticed the tobacco industry's efforts to prevent local regulations. Working on behalf of firearm enthusiasts and producers, the National Rife Association (NRA) became a key advocate for state laws preempting gun control regulations \parencite{hicksHomeRuleBe2018}. Today, 43 states preempt at least some local regulation of guns and ammunition \parencite{giffordslawcenterPreemptionLocalLawsn.d.,schraggerStatePreemptionLocal2017}, with many of the laws dating back to the 1980s and 1990s. Of the eleven states that do not allow any firearm or ammunition ordinances, for instance, nine of them passed their preemption laws before 2000 \parencite{schraggerStatePreemptionLocal2017}. More recently, some states have adopted gun control preemption laws that impose severe penalties on non-compliant local officials and cities. While such penalties are not exclusively associated with laws preventing local firearm regulation, many of the penalties for not complying with gun preemption laws are especially severe. Kentucky criminalizes non-compliance with the state preemption law, while local officials in Florida can be fined and even removed from office for not complying the state's gun preemption law \parencite{briffaultChallengeNewPreemption2018}.

@@

\begin{figure}[!ht]
\caption{State preemption of economic and labor ordinances, 1984-2019}
\label{fig:policies}
\centering
\includegraphics[width=.75\textwidth]{plots/preemption_plot}
\newline\scriptsize{Note: FILL IN. Data collected from, and plot design based on, \textcite{economicpolicyinstituteWorkerRightsPreemption2018}.}
\end{figure}

@@

The first condition concerns the threat of local policymaking. The states govern their cities using a variety of legal regimes, resulting in variation in local autonomy and the likelihood of municipal policymaking. If institutional rules prevent cities from passing regulations in the first place, states should not be expected to take the time to preempt specific municipal ordinances. The second condition concerns the ideological divide between cities and their states. If cities and states possess different preferences over the given regulatory measure, we should expect states to be more willing to preempt local authority. In states where city and state preferences are closer to one another, states will be more willing to let localities engage in policymaking. Finally, the third condition concerns the activity of relevant interest groups. The narrative above suggests that tobacco lobbying firms, the NRA, and ALEC have played a key role in mobilizing states to clamp down on local policymaking. Where these groups are more powerful, we should see more preemption activity.

@@

I hope to show that these conditions should apply to all preemption activity, regardless of the policy area. While the broad applicability of the hypotheses is an advantage, the reality of the data on state preemption and local policymaking means that testing hypotheses across multiple policy domains remains a distant goal. Collecting data on cities' policymaking (which is necessary to test whether the realization of the threat of local action spurs states to act preemptively) is no small task. I only know of one resource, the U.S. Tobacco Control Laws Database (TCLD) maintained by the American Nonsmokers' Rights Foundation,\footnote{\url{https://no-smoke.org/materials-services/lists-maps/}} that tracks both local ordinances and state preemption laws. While collecting data on cities' attempts to pass other regulations is important (especially in the policy areas that have seen the sharpest rise in preemption activity in recent years), my aims are more modest. I hope to use the TCLD's data on local tobacco regulation and state preemption over the past 40 years to test whether my three conditions hold. In the following pages, I provide further justification of the three conditions and offer concrete means of testing them in the context of tobacco preemption.

\subsubsection*{Hypothesis 1}
The first condition concerns cities' capacity to pass regulations. As described above, cities have often lacked this power. Early formulations of Home Rule, paired with narrow interpretations of cities' sphere of policymaking authority, hindered cities' ability to legislate as they saw fit. Since mid-century, many states have adopted legislative Home Rule regimes that permit a greater degree of state policymaking while allowing state courts and legislatures to intervene and deem local ordinances in conflict with state law. Nevertheless, there still remains considerable variation in the legal regimes governing city-state relations. This patchwork of legal rules results in differences across states in the likelihood of state preemption. In states where Dillon's Rule is still frequently applied by judges, cities possess limited policymaking authority. They can only legislate in policy domains that have been explicitly permitted by the state. While states can theoretically grant wide authority to cities and then preempt ordinances by curtailing city's delegated power, states that employ Dillon's Rule usually delegate narrow powers to cities in the first place. As a result, cities have little capacity to legislate and if they do act outside the scope of their delegated authority, it usually falls to the state courts to determine if the ordinance is legal \parencite{dillerIntrastatePreemption2007}.

In other states, cities' authority is structured according to the earlier formulation of Home Rule. Cities enjoy authority over matters of ``local concern,'' which is left open to judges' interpretation. Consequently, if a city ordinance is preempted, it is usually the result of a court deciding that the city acted outside of its local area of concern \parencite{dillerIntrastatePreemption2007}. In contrast, legislative Home Rule empowers cities to pass any ordinance that does not conflict with state law. While cities enjoy broad authority in such a regime, state legislatures, as discussed above, also possess the authority to intervene and curtail the bounds of local autonomy. Therefore, it is in states that govern cities with legislative Home Rule that we should expect to see the greatest level of preemption activity.

Unfortunately, given the complexities of state-local relations, there is no clear method for classifying states' legal regimes. Even the distinction between Home Rule and Dillon's Rule has been challenged \parencite{richardsonDillonRuleMars2011}. In a review of different classification schemes, \textcite{dillerIntrastatePreemption2007} notes that some scholars count 48 states with Home Rule while others count 45. Differentiating between the type of Home Rule is even trickier. One study counts 26 states with legislative Home Rule while others argue that the vast majority of Home Rule states employ legislative Home Rule. Moreover, classification disagreements do not only occur at the margin. In their analysis of the causes of state preemption, \textcite{fowlerStatePreemptionLocal2019}, citing \textcite{kraneHomeRuleAmerica2000}, only treat 10 states as using Home Rule. Deciding upon a classification scheme will be a necessary but challenging step as this project progresses.

While states' legal regimes provide one perspective on the threat of local action, another more concrete measure is whether a city in a state has already passed a tobacco regulation. Unlike the legal rules governing cities, which measures the potential for local action, whether a city has already regulated the sale or consumption of tobacco captures the realization of the threat. Controlling for cities' policymaking is notably absent from recent quantitative analyses of preemption \parencite{fowlerStatePreemptionLocal2019,goodmanStateLegislativeIdeology2019,flavinExplainingStatePreemption2019}, likely due to the difficulty of collecting data on cities' policies. Luckily, the TCLD systematically tracks local tobacco regulations alongside states' preemption laws. It includes the tobacco regulations of 98\% of the cities in the U.S. with more than 75,000 inhabitants, allowing me to test my hypotheses across all fifty states.

\subsubsection*{Hypothesis 2}
The second condition contends that diverging policy preferences between the state government and at least one city in the state increases the likelihood of state preemption. Struggles between cities and states over the bounds of local autonomy usually pit conservative state governments against liberal cities \parencite{rapoportBlueCitiesBattle2016,riverstone-newellRiseStatePreemption2017}, although this is not always the case.\footnote{For instance, in 2013, Oregon preempted cities from regulating genetically-modified crops and, in 2017, New York overturned New York City's tax on plastic bags \parencite{mckinleyCuomoBlocksNew2017,scharffHyperPreemptionReordering2018}.} Given the increasing number of state preemption laws targeting local economic regulations and increasing polarization between Republicans and Democrats on economic and labor issues over the past two decades \parencite{grumbachBackwatersMajorPolicymakers2018}, Republican control of government and state conservatism play an unsurprisingly key explanatory role in recent studies of preemption \parencite{hicksHomeRuleBe2018,riverstone-newellRiseStatePreemption2017,flavinExplainingStatePreemption2019,fowlerStatePreemptionLocal2019}.

While Republican control of government and conservatism certainly correlate with recent preemption activity, we do not yet have a clear idea of whether partisanship and ideology matter in absolute or relative terms. In other words, we do not know whether Republican control of government and conservatism inherently lead to preemption or whether the connection hinges on the relative difference between the preferences and ideology of the state and its cities. Should we expect preemption to be just as likely in North Dakota as in Texas, since they are both conservative states controlled by Republicans? Or does the fact that Texas contains much more progressive municipalities mean that it is more likely to preempt local ordinances? Based on the above discussion of preemption's history, where preemption often seems to occur as a response to local actions, the latter framing seems to be a more precise formulation of the role of partisanship and ideology. It is also more generalizable, as it does not presume that certain parties and ideologies always pursue preemption.

While treating ideology and policy preferences in relative terms is important, measuring the liberal ideology of cities is difficult, especially over multiple decades. One, albeit imperfect, approach is to use the share of votes earned by Democratic presidential candidates in urban counties. Figure \ref{fig:ideology} uses this approach to display the growing ideological divide between states and cities. The dashed lines in the figure show the share of votes for the Democratic presidential candidate in the three most populous counties in a state divided by the candidate's vote share in the entire state, averaged over all states that supported the Democratic and Republican candidates, respectively. In 1980, the most populous counties in states that supported either the Republican or Democratic candidate supported the Democratic candidate to roughly the same degree as the entire state. Beginning in the mid-1980s, the share of votes dedicated to the Democratic candidate in urban counties began to increase relative to the Democratic candidate's share of votes in the entire state. The divergence is especially pronounced in states that supported the Republican candidate in the election, implying that urban counties in red states are becoming increasingly liberal relative to their states.

@@

The above discussion suggests that we need to move beyond treating Republican control of government and conservative ideology as intrinsically linked to state preemption of local laws. While increasing polarization and diverging preferences between rural and urban areas means that conservative, Republican-controlled states are often the source of recent preemption activity, it does not seem that this has always been the case. The ideological and partisan geography of the U.S. in the late 1980s and 1990s suggests that preemption must have been occurring in Democratic states. It will take getting the TCLD data in-hand for me to determine whether this circumstantial evidence holds up to further scrutiny.
